---
title: "Retrieve and Use Mass Spectrometry Data from MassIVE"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Retrieve and Use Mass Spectrometry Data from MassIVE}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{MsBackendMassIVE}
    %\VignetteDepends{Spectra,BiocStyle}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendMassIVE")`<br />
**Authors**: `r packageDescription("MsBackendMassIVE")[["Author"]] `<br />
**Last modified:** `r file.info("MsBackendMassIVE.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(BiocStyle)
```

# Introduction

The `r Biocpkg("Spectra")` package provides a central infrastructure for the
handling of Mass Spectrometry (MS) data in Bioconductor. The package supports
interchangeable use of different *backends* to import and represent MS data from
a variety of sources and data formats. The *MsBackendMassIVE* package allows to
retrieve MS data files directly from the
[MassIVE](https://massive.ucsd.edu/ProteoSAFe/static/massive.jsp) (Mass
Spectrometry Interactive Virtual Environment) repository.
MassIVE is a community resource developed by the NIH-funded Center for
Computational Mass Spectrometry at UC San Diego to promote the global, free
exchange of mass spectrometry data. MassIVE supports deposition of both
proteomics and metabolomics experiments and is a full member of the
[ProteomeXchange](http://www.proteomexchange.org/) consortium.

The *MsBackendMassIVE* package downloads and locally caches MS data files for a
MassIVE data set and enables further analyses of this data directly in R.


# Installation

The package can be installed from within R with the commands below:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("RforMassSpectrometry/MsBackendMassIVE")
```


# Importing MS Data from MassIVE

Each experiment in MassIVE is identified by a unique accession starting with
*MSV* followed by a number. The MS data files of a dataset are listed in the
[GNPS2 database](https://datasetcache.gnps2.org/datasette/database/filename)
and can be downloaded from MassIVE's FTP server.

Below we list all files from the MassIVE data set with the ID *MSV000080547*.

```{r}
library(MsBackendMassIVE)

#' List files of a MassIVE data set
all_files <- massive_list_files("MSV000080547")
head(all_files)
```

These files are accessible through the FTP path associated with the MassIVE
data set. Below we use the `massive_ftp_path()` function to return the FTP path
for our test data set.

```{r}
massive_ftp_path("MSV000080547", mustWork = FALSE)
```

```{r, echo = FALSE}
Sys.sleep(4)
```

MS data files in supported formats (mzML, CDF, mzXML) can be directly loaded
using the `MsBackendMassIVE` backend into R as a `Spectra` object
(`MsBackendMassIVE` directly extends *Spectra*'s `MsBackendMzR` backend and
therefore supports import of MS data files in *mzML*, *CDF* or *mzXML*
formats). By default, all MS data files of the data set would be retrieved, but
in our example below we restrict to a few data files to reduce the amount of
data that needs to be downloaded. To this end we define a pattern matching the
file name of only some data files using the `filePattern` parameter.

```{r}
library(Spectra)

#' Load MS data files of one data set
s <- Spectra("MSV000080547", filePattern = "1.mzML$",
             source = MsBackendMassIVE())
s
```

This call downloaded the files to the local cache and loaded them as a `Spectra`
object. The downloading and caching of the data is handled by Bioconductor's
`r Biocpkg("BiocFileCache")`. The local cache can thus be managed directly using
functionality from that package. Any subsequent loading of the same data files
will load the locally cached versions avoiding thus repetitive download of the
same data.

The `Spectra` object with the MS data files of the MassIVE data set enables now
any subsequent analysis of the data in R. On top of the spectra variables and
mass peak data values that are provided by the MS data files also additional
information related to the MassIVE data set are available as specific *spectra
variables*. We list all available spectra variables of the data set below.

```{r}
spectraVariables(s)
```

The MassIVE-specific variables are `"massive_id"` and `"data_file"` providing
the MassIVE ID of the data set and the original data file name on the MassIVE
FTP server for each individual spectrum.

```{r}
spectraData(s, c("massive_id", "data_file"))
```

The `massive_sync()` function can be used to *synchronize* the local content of
a `MsBackendMassIVE`. This function checks if all data files of the backend are
available locally and eventually downloads and caches missing files.

```{r, echo = FALSE}
Sys.sleep(4)
```

```{r}
massive_sync(s@backend)
```

Also, it is possible to *manually* cache and download data files from MassIVE
using the `massive_sync_data_files()` function. This function evaluates if the
respective data files are already cached and, if so, does not download them
again. Below we use this to retrieve the local storage information on one of the
data files of the MassIVE data set *MSV000080547*:

```{r}
res <- massive_sync_data_files("MSV000080547",
                               fileName = "AG_spiked_sample11.mzML")
res
```

The `massive_cached_data_files()` function can be used to inspect and list
locally cached MassIVE data files. This function does not require an active
internet connection since only local content is queried. With the default
settings, a `data.frame` with all available data files is returned.

```{r}
massive_cached_data_files()
```

Locally cached files for a MassIVE data set can be removed using the
`massive_delete_cache()` function providing the ID of the MassIVE data set for
which local data files should be removed.

# Session information

```{r}
sessionInfo()
```
